Minimal trap spaces of Logical models are maximal siphons of their Petri net encoding

### 2022-05-10 cmsb22.tex
Updates from Overleaf
yalafi (languagetool)

Petri nets were introduced in the 60’s as simple formalism for describing and analyzing information processing systems that are characterized as being concurrent, asynchronous, non-deterministic and possibly distributed~\cite{peterson1981petri,Murata1989}.
The use of Petri nets for representing biochemical reaction systems, by mapping molecular species to places and reactions to transitions, was introduced quite late in~\cite{reddy1993petri}, together with some Petri net concepts and tools for the analysis of metabolic networks.
Siphons are such a concept but they have not been used a lot for the analysis of biochemical systems even if the practical cost of computing their minimal/maximal elements appear much more manageable than the theoretical complexity would indicate~\cite{nabli2016enumerating}.

@@

\end{definition}
A \emph{marking} for a Petri net is a mapping \(m : P \mapsto \mathbb{N}\) that assigns a number of tokens to each place. A place \(p\) is marked by a marking \(m\) if and only if \(m(p) > 0\). We shall write \(pred(x)\) (resp.\ \(succ(x)\)) to represent the set of vertices that have an (non-zero weighted) arc leading to (resp.\ coming from) \(x\).

The link between Logical models \emph{à la} Thomas and Petri nets was originally established in~\cite{chaouiya2004qualitative} in order to make available formal methods like model-checking for the analysis of such systems.
The basic encoding into 1-safe (i.e., never more than one token in each place) nets only holds for purely Boolean models but was later extended to multi-valued models in two ways, either in \cite{chaouiya2011petri} with non 1-safe Petri nets or more recently in~\cite{chatain2014characterization} with 1-safe nets but many more places.

Since our study is focused on Boolean models, we briefly recall the original encoding here.
Its basis is that every node (\emph{gene}) \(v\) of the original model \(\mathcal{M} = (V, F)\) is represented by two separate places (\(p_v\) and \(\overline{p}_v\)), corresponding to its two states, active, and inactive, respectively.
Each conjunct of the logical function that activates the \emph{gene} will lead to a transition \(t\), consuming the inactive place (i.e., a directional arc from \(\overline{p}_v\) to \(t\)), producing the active place (i.e., a directional arc from \(t\) to \(p_v\)), and with all other literals both consumed and produced (i.e., a bi-directional arc).
And conversely for the inactivation. 
Let \(s\) be a state of the Boolean model and \(m_s\) be its the corresponding marking in the encoded Petri net. It holds that \(\forall v \in V\), \(s(v) = 0\) if and only if \(m_s(\overline{p}_v) = 1\) and \(s(v) = 1\) if and only if \(m_s(p_v) = 1\). Note also that at any marking \(m\) of the Petri net encoding a Boolean model, it always holds that \(m(p_v) + m(\overline{p}_v) = 1\).

@@

  First, we show that if \(m\) is a trap space of \(\mathcal{M}\), then \(S\) is a conflict-free siphon of \(\mathcal{P}\) (*). If \(D_m = \emptyset\), then \(S = \emptyset\) is trivially a conflict-free siphon of \(\mathcal{P}\). Thus, we consider the case that \(D_m \neq \emptyset\) (resp. \(S \neq \emptyset\)). Assume that \(S\) is not a siphon of \(\mathcal{P}\). Then, there is a transition \(t \in T\) such that \(S\cap succ(t)\not =\emptyset\) but \(S\cap pred(t)=\emptyset\). In other words, there is a place \(p \in S\) such that \(p \in succ(t)\) but \(p \not \in pred(t)\). Let \(v\) be the corresponding node in \(\mathcal{M}\) of \(p\). By the characterization of the encoding~\cite{chaouiya2004qualitative}, there is a directional arc from \(t\) to \(p\) and a directional arc from the complementary place of \(p\) to \(t\). Without loss of generality, we assume that \(p = p_v\), then there is a directional arc from \(t\) to \(p_v\) and a directional arc from \(\overline{p}_v\) to \(t\). In addition, there is also no arc or a bi-directional arc between \(t\) and another place rather than \(p_v\) and \(\overline{p}_v\). Thus, there is no connecting arc between \(t\) and any place in \(S \setminus \{p_v\}\) because \(S\cap pred(t)\not =\emptyset\). In \(\mathcal{S}_{\mathcal{M}}[m]\), a node in \(V \setminus D_m\) can receive any Boolean value. Hence, there is a state \(s \in \mathcal{S}_{\mathcal{M}}[m]\) such that \(m_s(p') = 1, \forall p' \in pred(t) \setminus \{\overline{p_v}\}\) where \(m_s\) is the corresponding marking in \(\mathcal{P}\) of \(s\). We also have \(m_s(p_v) = 0\), leading to \(m_s(\overline{p_v}) = 1\) by the characterization of the encoding~\cite{chaouiya2004qualitative}. Now, \(t\) is enabled at marking \(m_s\). Its firing leads to a new marking \(m'_s\) such that \(m'_s(p_v) = 1\) and \(m'_s(\overline{p}_v) = 0\). Let \(s'\) be the corresponding state in \(\mathcal{M}\) of \(m'_s\). Since \(m\) is a trap space of \(\mathcal{M}\), \(s' \in \mathcal{S}_{\mathcal{M}}[m]\). Then, \(s'(v) = m(v)\), leading to \(m'_s(p_v) = 0\), which is a contradiction. Hence, \(S\) is a siphon of \(\mathcal{P}\). By the definition of a mirror, \(S\) is also a conflict-free one.

@@

\section{Answer set programming-based method}
First, we show the characterization of all conflict-free siphons of the encoded Petri net \(\mathcal{P} = (P, T, W)\). Suppose that \(S\) is a generic siphon of \(\mathcal{P}\). If a place \(p\) should belong to \(S\), then by definition all the transitions in \(pred(p)\) must belong to \(succ(S)\). Note that \(succ(S) = \bigcup_{p \in S}succ(p)\). A transition \(t\) belongs to \(succ(S)\) if and only if there is at least one place \(p'\) in \(S\) such that \(p' \in pred(t)\). Hence, for each transition \(t \in pred(p)\), we can state that
\begin{equation}
\label{eq:siphon}
p \in S \Rightarrow \bigvee_{p' \in pred(t)}p' \in S.
\end{equation}The system of all the rules of the above form with respect to all pairs \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\) fully characterizes all generic siphons of a Petri net~\cite{nabli2016enumerating}.
To make \(S\) to be a conflict-free siphon, we need to add to the system the rule
\begin{equation}
\label{eq:conflict}
p_v \in S \Rightarrow \overline{p_v} \not \in S \wedge \overline{p_v} \in S \Rightarrow p_v \not \in S
\end{equation}for each node \(v \in V\).
By definition, the final system fully characterizes all conflict-free siphons of the encoded Petri net.

@@

In practice, even if there is a huge number of attractors in such a model, obtaining a sample of those can reveal very useful to invalidate the model and lead to further refinement.
In particular, it provides a feature-rich alternative to random simulations for this type of very non-deterministic model.
Being able to detect that there are inconsistencies with published experimental data in some of the, say first 100 attractors, can lead to a much quicker Systems Biology loop: model, invalidate, refine.

@@

To assess the efficiency of the proposed method, we compared it with the state-of-the-art method PyBoolNet~\cite{klarner2017pyboolnet}.
We used a set of real-world Boolean models lying in various scales collected from various bibliographic sources.
These models are quite big (in size), complex (i.e., having high average in-degree, which is related to the number of prime-implicants) and most of them have never been fully analyzed.
We then applied the proposed method and PyBoolNet to computing minimal trap spaces of these real-world models.
It is notable that unlike existing analysis shown in the literature, we did not fix specific values for source nodes (i.e., some node \(v\) such that \(f_v = v\)) in these models.
To solve the ASP problems, we used the same ASP solver CLINGO~\cite{DBLP:journals/aicom/GebserKKOSS11} and the same configuration as that used in PyBoolNet~\cite{klarner2015computing}.
Specifically, we used the configuration \texttt{-heuristic=Domain -enum-mod=domRec -dom-mod=3} (subset maximality, equivalent to the deprecated \texttt{--dom-pref=32 --heu=domain --dom-mod=7} used by PyBoolNet). We ran all the benchmarks on a virtual machine whose environment is CPU: Intel(R) Core(TM) i7-3630QM 2.40GHz x 4, Memory: 8 GB, Ubuntu 18.04.2 64 bit. Finally, we set a time limit as one hour for each model.

>>>>

Petri nets were introduced in the 60s as a simple formalism for describing and analyzing information processing systems that are characterized as being concurrent, asynchronous, non-deterministic and possibly distributed~\cite{peterson1981petri,Murata1989}.
The use of Petri nets for representing biochemical reaction systems, by mapping molecular species to places and reactions to transitions, was introduced quite late in~\cite{reddy1993petri}, together with some Petri net concepts and tools for the analysis of metabolic networks.
Siphons are such a concept, but they have not been used a lot for the analysis of biochemical systems even if the practical cost of computing their minimal/maximal elements appear much more manageable than the theoretical complexity would indicate~\cite{nabli2016enumerating}.

@@

\end{definition}
A \emph{marking} for a Petri net is a mapping \(m : P \mapsto \mathbb{N}\) that assigns a number of tokens to each place. A place \(p\) is marked by a marking \(m\) if and only if \(m(p) > 0\). We shall write \(pred(x)\) (resp.\ \(succ(x)\)) to represent the set of vertices that have a (non-zero weighted) arc leading to (resp.\ coming from) \(x\).

The link between Logical models \emph{à la} Thomas and Petri nets was originally established in~\cite{chaouiya2004qualitative} in order to make available formal methods like model-checking for the analysis of such systems.
The basic encoding into 1-safe (i.e., never more than one token in each place) nets only holds for purely Boolean models but was later extended to multivalued models in two ways, either in~\cite{chaouiya2011petri} with non 1-safe Petri nets or more recently in~\cite{chatain2014characterization} with 1-safe nets but many more places.

Since our study is focused on Boolean models, we briefly recall the original encoding here.
Its basis is that every node (\emph{gene}) \(v\) of the original model \(\mathcal{M} = (V, F)\) is represented by two separate places (\(p_v\) and \(\overline{p}_v\)), corresponding to its two states, active, and inactive, respectively.
Each conjunct of the logical function that activates the \emph{gene} will lead to a transition \(t\), consuming the inactive place (i.e., a directional arc from \(\overline{p}_v\) to \(t\)), producing the active place (i.e., a directional arc from \(t\) to \(p_v\)), and with all other literals both consumed and produced (i.e., a bidirectional arc).
And conversely for the inactivation. 
Let \(s\) be a state of the Boolean model and \(m_s\) be its corresponding marking in the encoded Petri net. It holds that \(\forall v \in V\), \(s(v) = 0\) if and only if \(m_s(\overline{p}_v) = 1\) and \(s(v) = 1\) if and only if \(m_s(p_v) = 1\). Note also that at any marking \(m\) of the Petri net encoding a Boolean model, it always holds that \(m(p_v) + m(\overline{p}_v) = 1\).

@@

  First, we show that if \(m\) is a trap space of \(\mathcal{M}\), then \(S\) is a conflict-free siphon of \(\mathcal{P}\) (*). If \(D_m = \emptyset\), then \(S = \emptyset\) is trivially a conflict-free siphon of \(\mathcal{P}\). Thus, we consider the case that \(D_m \neq \emptyset\) (resp. \(S \neq \emptyset\)). Assume that \(S\) is not a siphon of \(\mathcal{P}\). Then, there is a transition \(t \in T\) such that \(S\cap succ(t)\not =\emptyset\) but \(S\cap pred(t)=\emptyset\). In other words, there is a place \(p \in S\) such that \(p \in succ(t)\) but \(p \not \in pred(t)\). Let \(v\) be the corresponding node in \(\mathcal{M}\) of \(p\). By the characterization of the encoding~\cite{chaouiya2004qualitative}, there is a directional arc from \(t\) to \(p\) and a directional arc from the complementary place of \(p\) to \(t\). Without loss of generality, we assume that \(p = p_v\), then there is a directional arc from \(t\) to \(p_v\) and a directional arc from \(\overline{p}_v\) to \(t\). In addition, there is also no arc or a bidirectional arc between \(t\) and another place rather than \(p_v\) and \(\overline{p}_v\). Thus, there is no connecting arc between \(t\) and any place in \(S \setminus \{p_v\}\) because \(S\cap pred(t)\not =\emptyset\). In \(\mathcal{S}_{\mathcal{M}}[m]\), a node in \(V \setminus D_m\) can receive any Boolean value. Hence, there is a state \(s \in \mathcal{S}_{\mathcal{M}}[m]\) such that \(m_s(p') = 1, \forall p' \in pred(t) \setminus \{\overline{p_v}\}\) where \(m_s\) is the corresponding marking in \(\mathcal{P}\) of \(s\). We also have \(m_s(p_v) = 0\), leading to \(m_s(\overline{p_v}) = 1\) by the characterization of the encoding~\cite{chaouiya2004qualitative}. Now, \(t\) is enabled at marking \(m_s\). Its firing leads to a new marking \(m'_s\) such that \(m'_s(p_v) = 1\) and \(m'_s(\overline{p}_v) = 0\). Let \(s'\) be the corresponding state in \(\mathcal{M}\) of \(m'_s\). Since \(m\) is a trap space of \(\mathcal{M}\), \(s' \in \mathcal{S}_{\mathcal{M}}[m]\). Then, \(s'(v) = m(v)\), leading to \(m'_s(p_v) = 0\), which is a contradiction. Hence, \(S\) is a siphon of \(\mathcal{P}\). By the definition of a mirror, \(S\) is also a conflict-free one.

@@

\section{Answer set programming-based method}
First, we show the characterization of all conflict-free siphons of the encoded Petri net \(\mathcal{P} = (P, T, W)\). Suppose that \(S\) is a generic siphon of \(\mathcal{P}\). If a place \(p\) should belong to \(S\), then by definition all the transitions in \(pred(p)\) must belong to \(succ(S)\). Note that \(succ(S) = \bigcup_{p \in S}succ(p)\). A transition \(t\) belongs to \(succ(S)\) if and only if there is at least one place \(p'\) in \(S\) such that \(p' \in pred(t)\). Hence, for each transition \(t \in pred(p)\), we can state that
\begin{equation}
\label{eq:siphon}
p \in S \Rightarrow \bigvee_{p' \in pred(t)}p' \in S.
\end{equation}
The system of all the rules of the above form with respect to all pairs \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\) fully characterizes all generic siphons of a Petri net~\cite{nabli2016enumerating}.
To make \(S\) to be a conflict-free siphon, we need to add to the system the rule
\begin{equation}
\label{eq:conflict}
p_v \in S \Rightarrow \overline{p_v} \not \in S \wedge \overline{p_v} \in S \Rightarrow p_v \not \in S
\end{equation}for each node \(v \in V\).
By definition, the final system fully characterizes all conflict-free siphons of the encoded Petri net.

@@

In practice, even if there is a huge number of attractors in such a model, obtaining a sample of those can reveal very useful to invalidate the model and lead to further refinement.
In particular, it provides a feature-rich alternative to random simulations for this type of very non-deterministic model.
Being able to detect that there are inconsistencies with published experimental data in some of the first 1000 attractors, for instance, can lead to a much quicker Systems Biology loop: model, invalidate, refine.

@@

To assess the efficiency of the proposed method, we compared it with the state-of-the-art method PyBoolNet~\cite{klarner2017pyboolnet}.
We used a set of real-world Boolean models lying in various scales collected from various bibliographic sources.
These models are quite big (in size), complex (i.e., having high average in-degree, which is related to the number of prime-implicants) and most of them have never been fully analyzed.
We then applied the proposed method and PyBoolNet to computing minimal trap spaces of these real-world models.
It is notable that unlike existing analysis shown in the literature, we did not fix specific values for source nodes (i.e., some node \(v\) such that \(f_v = v\)) in these models.
To solve the ASP problems, we used the same ASP solver CLINGO~\cite{DBLP:journals/aicom/GebserKKOSS11} and the same configuration as that used in PyBoolNet~\cite{klarner2015computing}.
Specifically, we used the configuration \texttt{-heuristic=Domain -enum-mod=domRec -dom-mod=3} (subset maximality, equivalent to the deprecated \texttt{--dom-pref=32 --heu=domain --dom-mod=7} used by PyBoolNet). We ran all the benchmarks on a virtual machine whose environment is CPU: Intel® Core™ i7-3630QM 2.40GHz x 4, Memory: 8 GB, Ubuntu 18.04.2 64 bit. Finally, we set a time limit as one hour for each model.
